Manuscript Figures

### 2021-03-22 src/manuscript_figures.Rmd
simplify cluster stats, update figures for msi stats

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
# todo: pass this as argument
settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas'

@@

```{r import_lfs}
LFs <- sapply(simplify = F, TOOLS, function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'factors.csv$', full.names = T)
  LFs <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- dt$V1
    return(M)
  })
  names(LFs) <- unlist(lapply(strsplit(basename(names(LFs)), split = '\\.'), function(x) paste(x[1:2], collapse = '.')))
  return(LFs)
})
```

```{r import_feature_weights}
FWs <- sapply(simplify = F, TOOLS, function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'feature_weights.csv$', full.names = T)
  W <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- gsub("^assay: ", "", dt$V1) # required only for maui
    return(M)
  })
  names(W) <- unlist(lapply(strsplit(basename(names(W)), split = '\\.'), 
                            function(x) paste(x[1:2], collapse = '.')))
  return(W)
})
```

@@

multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
clusters <- sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      M <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
    } 
    M <- t(subset_features_by_variance(t(M), topN = 10))

    val <- c('internal') 
    if(ncol(M) > 2) {val <- c('internal', 'stability')}
    res <- clValid::clValid(obj = M,
                          nClust = 3:8,
                          clMethods = 'kmeans',
                          validation = val,
                          maxitems = 10000,
                          nstart = 1000)
    memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']])))
    colnames(memberships) <- paste0('k', colnames(memberships))
    stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
    colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
    return(list('clustering_stats' = stats, 'cluster_memberships' = memberships))
  })
})
parallel::stopCluster(cl)

cluster.stats <- do.call(rbind, lapply(names(clusters), function(tool) {
  do.call(rbind, lapply(names(clusters[[tool]]), function(experiment) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      datatypes <- strsplit(experiment, "\\.")[[1]][2]
      stats <- clusters[[tool]][[experiment]][['clustering_stats']]
      stats$cancer_type <- cancer_type
      stats$datatypes <- datatypes
      stats$method <- tool
      return(stats)
  }))
}))

# plot silhouette/stability score comparison
dt1 <- cluster.stats[metric == 'Silhouette'][datatypes == multi_omics_comb]
dt1$method <- toupper(gsub("_factors", "", dt1$method))
comps <- lapply(setdiff(unique(dt1$method), 'MAUI'), function(x) c('MAUI', x))

p1 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = comps) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')

dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb]
dt2$method <- toupper(gsub("_factors", "", dt2$method))
comps <- lapply(setdiff(unique(dt2$method), 'MAUI'), function(x) c('MAUI', x))
p2 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = comps) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')

# combine both scores and plot for different values of k
dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
            cancer_type + method + k ~ metric, value.var = 'score')
dt$kv <- paste0('kmeans(k=',dt$k,')')
dt$method <- toupper(gsub("_factors", "", dt$method))
p3 <- ggplot(dt,
       aes(x = Silhouette, y = APN)) + 
        geom_point(aes(color = method), size = 3) + 
        facet_wrap(~ kv, ncol = 2) +
        labs(title = 'Multi-omics clustering: Silhouette vs APN')

@@

dt <- melt(surv.stats, measure.vars = c('cindex.pfi', 'cindex.os'))
dt[is.na(value)]$value <- min(dt$value, na.rm = T) # 
dt$method <- toupper(gsub("_factors", "", dt$method))
dt$project <- unlist(sapply(strsplit(dt$experiment, "\\."), function(x) x[1]))
dt$endpoint <- ifelse(dt$variable == 'cindex.pfi', 'Progression Free Survival', 
                      'Overall Survival')
# PCA/MOFA win, compare against PCA
comps <- lapply(setdiff(unique(dt$method), 'PCA'), function(x) c('PCA', x))
p4 <- ggboxplot(dt, x = 'method', y = 'value', color = 'method',
                add = 'jitter', facet.by = 'endpoint') +
  stat_compare_means(method = 'wilcox.test', paired = TRUE,
                     comparisons = comps) +
  labs(y = "Harrel's C-index", x = '') +
  theme(legend.position = 'none')

Fig <- plot_grid(plot_grid(p1, p2, labels = c('A', 'B'), ncol = 1),
                 plot_grid(p3, p4, labels = c('C', 'D'), ncol = 1), 
                 ncol = 2, rel_widths = c(2,3))

@@

```{r msi_stats, fig.width=10, fig.height=8}
# Predicting MSI status using latent factors
# predict MSI status for pan-gastrointestinal tumors
# read MSI data
dataDir <- settings$GDCdataDir
MSI <- do.call(rbind, lapply(projectDict$`pan-gastrointestinal`, function(pr) {
  f <- file.path(dataDir, paste0(pr, '.MSI.tsv'))
  dt <- data.table::fread(f)[,c(2,4)]
  colnames(dt)[2] <- 'MSI'
  dt$project <- pr
  return(dt)
}))
# remove samples with undertermined MSI status
MSI <- MSI[MSI %in% c('MSI-H', 'MSI-L', 'MSS')]
msi_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
  y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y, partition = 0.6, num.trees = 2000)
  res <- predict_cluster_labels_glmnet(M, y)
  #res <- predict_labels_glmnet(M, y)
  res$method <- tool
  return(res)
}))
msi_stats$method <- toupper(gsub('_factors', '', msi_stats$method))

p1 <- ggplot(msi_stats[method == 'MAUI'], aes(x = label, y = auc)) +
  geom_bar(stat = 'identity', aes(fill = label)) + 
  labs(y = 'AUC', x = '') +
  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2)+
  theme(legend.title = element_blank())

# make some maui specific plots to explore MSI related LFs
M <- remove_redundant_variables(LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']], 
                                perc = 99) # make it consistent with pan-cancer
y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI 
M <- M[!is.na(y),]
y <- y[!is.na(y)]

# tsne plot of top LF related to MSI-H and MSI classes
dt <- get_factor_specific_variables(M, y)
# use top predictive lfs to guide tsne 
top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 5)[['variable']])

@@

p2.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
  theme(legend.title = element_blank())
p2.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
  geom_point(aes_string(color = top_lf)) + 
  scale_color_gradient(low = 'gray', high = 'red')

@@

# get top 100 features
df <- df[order(abs(df[,top_lf]), decreasing = T),]
freq <- do.call(rbind, lapply(0:29, function(i) {
  s <- (i*100+1)
  e <- (i+1)*100
  x <- data.frame(table(df[s:e,]$data.type))
  x$rank <- stringr::str_wrap(paste0("[",s,",",e,"]"),
                              width = 8)
  x$ind <- i
  return(x)
}))
p3 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
  geom_bar(aes(fill = Var1), stat='identity') + 
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45)) + 
  labs(x = paste0("top features contributing to ",top_lf))

Fig <- cowplot::plot_grid(p1, 
                          plot_grid(p2.1, p2.2, p3, nrow = 1, labels = c('B', 'C', 'D')),
                          nrow = 2, labels = c('A', ''), rel_heights = c(5,3))
fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.MSI.status.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', width = 10, height = 8)
Legends[[fig_count]] <- "Predicting MSI status of pan-gastrointestinal tumors using MAUI latent factors. A) AUC values for predicting MSI status using MAUI LFs B) Top features that contribute to MAUI latent factor (‘LF95’) that is the most predictive latent factor in distinguishing samples with high microsatellite instability from the rest. C) tSNE plots of samples generated using top MSI-predictive LFs (5 LF per MSI class) colored by MSI status D) tSNE plots of samples colored by MAUI latent factor (‘LF95’) which is most predictive of MSI-H samples."
```

@@

res <- predict_cluster_labels_glmnet(M, ct)

@@

Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
# todo: pass this as argument
settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas'

@@

```{r import_lfs}
LFs <- import_LFs(TOOLS, inputDir)
```

```{r import_feature_weights}
FWs <- import_feature_weights(TOOLS, inputDir)
```

@@

multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
cluster.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  stats <- do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, 
                                            intersect(names(LFs[[tool]]), experiments), 
                    function(experiment) {
    message(experiment)
    source(settings$utility_script)
    library(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      M <- remove_redundant_variables(M, perc = 99) 
    }

    # try topN from 5 to 25 (each tool may capture different amount of information 
    # by varying number of factors)
    stats <- do.call(rbind, lapply(seq(5, min(25, ncol(M)), 5), function(n) {
      message(n)
      M.sub <- t(subset_features_by_variance(t(M), topN = n))
      val <- c('internal') 
      if(ncol(M) > 2) {val <- c('internal', 'stability')}
      res <- clValid::clValid(obj = M.sub,
                            nClust = 3:8,
                            clMethods = 'kmeans',
                            validation = val,
                            maxitems = 10000,
                            nstart = 1000, iter.max = 30)
      stats <- data.table::as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
      colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
      stats$top_factors <- n
      return(stats)
    }))
    stats$cancer_type <- strsplit(experiment, "\\.")[[1]][1]
    stats$datatypes <- strsplit(experiment, "\\.")[[1]][2]
    return(stats)
  }))
  stats$method <- tool
  return(stats)
}))
parallel::stopCluster(cl)

# plot silhouette/stability score comparison
dt1 <- cluster.stats[metric == 'Silhouette'][datatypes == multi_omics_comb,
                                             .SD[which.max(score)],
                                             by = c('cancer_type', 'method', 'k')]
dt1$method <- toupper(gsub("_factors", "", dt1$method))
comps <- lapply(setdiff(unique(dt1$method), 'MAUI'), function(x) c('MAUI', x))

p1 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = FALSE, #todo change to true when mofa is finished
                     comparisons = comps) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')

dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb,
                                             .SD[which.min(score)],
                                             by = c('cancer_type', 'method', 'k')]
dt2$method <- toupper(gsub("_factors", "", dt2$method))
comps <- lapply(setdiff(unique(dt2$method), 'MAUI'), function(x) c('MAUI', x))
p2 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = FALSE,  #todo change to true when mofa is finished
                     comparisons = comps) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')

# combine both scores and plot for different values of k
# dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
#             cancer_type + method + k ~ metric, value.var = 'score')
# dt$kv <- paste0('kmeans(k=',dt$k,')')
# dt$method <- toupper(gsub("_factors", "", dt$method))
# p3 <- ggplot(dt,
#        aes(x = Silhouette, y = APN)) + 
#         geom_point(aes(color = method), size = 3) + 
#         facet_wrap(~ kv, ncol = 2) +
#         labs(title = 'Multi-omics clustering: Silhouette vs APN')

@@

dt <- melt(surv.stats, measure.vars = c('cindex.pfi', 'cindex.os'))
dt[is.na(value)]$value <- min(dt$value, na.rm = T) # 
dt$method <- toupper(gsub("_factors", "", dt$method))
dt$project <- unlist(sapply(strsplit(dt$experiment, "\\."), function(x) x[1]))
dt$endpoint <- ifelse(dt$variable == 'cindex.pfi', 'Progression Free Survival', 
                      'Overall Survival')
# PCA/MOFA win, compare against PCA
comps <- lapply(setdiff(unique(dt$method), 'MAUI'), function(x) c('MAUI', x))
p4 <- ggboxplot(dt, x = 'method', y = 'value', color = 'method',
                add = 'jitter', facet.by = 'endpoint') +
  stat_compare_means(method = 'wilcox.test', paired = TRUE,
                     comparisons = comps) +
  labs(y = "Harrel's C-index", x = '') +
  theme(legend.position = 'none')

Fig <- plot_grid(plot_grid(p1, p2, labels = c('A', 'B'), nrow = 1),
                 p4, nrow = 2, labels = c('', 'C'))

@@

```{r msi_stats, fig.width=10, fig.height=8}
# Predicting MSI status using latent factors
# predict MSI status for pan-gastrointestinal tumors
# read MSI data
dataDir <- settings$GDCdataDir
MSI <- do.call(rbind, lapply(projectDict$`pan-gastrointestinal`, function(pr) {
  f <- file.path(dataDir, paste0(pr, '.MSI.tsv'))
  dt <- data.table::fread(f)[,c(2,4)]
  colnames(dt)[2] <- 'MSI'
  dt$project <- pr
  return(dt)
}))
# remove samples with undertermined MSI status
MSI <- MSI[MSI %in% c('MSI-H', 'MSI-L', 'MSS')]

M <- remove_redundant_variables(LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']], 
                                perc = 99)
y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
M <- M[!is.na(y),]
y <- y[!is.na(y)]

res <- predict_cluster_labels_glmnet(M, y)
p1 <- ggplot(res, aes(x = label, y = auc)) +
  geom_bar(stat = 'identity', aes(fill = label)) + 
  labs(y = 'AUC', x = '') +
  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2)+
  theme(legend.title = element_blank())

# also plot a roc curve
rocs <- predict_cluster_labels_glmnet(M, y, returnROC = TRUE)
p2 <- plot_roc_list(rocs)

# explore MSI related LFs
# tsne plot of top LF related to MSI-H and MSI classes
dt <- get_factor_specific_variables(M, y)
# use top predictive lfs to guide tsne 
top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 5)[['variable']])

@@

p3.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
  theme(legend.title = element_blank())
p3.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
  geom_point(aes_string(color = top_lf)) + 
  scale_color_gradient(low = 'gray', high = 'red')

@@

# get top 100 features
df <- df[order(abs(df[,top_lf]), decreasing = T),]
freq <- do.call(rbind, lapply(0:29, function(i) {
  s <- (i*100+1)
  e <- (i+1)*100
  x <- data.frame(table(df[s:e,]$data.type))
  x$rank <- stringr::str_wrap(paste0("[",s,",",e,"]"),
                              width = 8)
  x$ind <- i
  return(x)
}))
p4 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
  geom_bar(aes(fill = Var1), stat='identity') + 
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45)) + 
  labs(x = paste0("top features contributing to ",top_lf))

Fig <- cowplot::plot_grid(plot_grid(p1, p2, ncol = 1, labels = c('A', 'B')),
                          plot_grid(p3.1, p3.2, p4, ncol = 1, labels = c('C', 'D', 'E')),
                          nrow = 1)
fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.MSI.status.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', width = 8, height = 8)
Legends[[fig_count]] <- "Predicting MSI status of pan-gastrointestinal tumors using MAUI latent factors. A) AUC values for predicting MSI status using MAUI LFs B) Top features that contribute to MAUI latent factor (‘LF95’) that is the most predictive latent factor in distinguishing samples with high microsatellite instability from the rest. C) tSNE plots of samples generated using top MSI-predictive LFs (5 LF per MSI class) colored by MSI status D) tSNE plots of samples colored by MAUI latent factor (‘LF95’) which is most predictive of MSI-H samples."
```

@@

@@

bipartite_plots <- unlist(lapply(1:(ncol(df)-1), function(i) {
  lapply((i+1):ncol(df), function(j) {
    p <- plot_cluster_comparison(df[,i,drop = F], df[,j,drop = F]) + 
      theme(plot.margin = unit(c(0.2, 1.5, 0, 1.5), units = 'in'))
  })
}), recursive = F)

Fig1 <- cowplot::plot_grid(plotlist = bipartite_plots, nrow = 3, labels = 'AUTO')
ggsave('test1.pdf', plot = Fig1, units = 'in', width = 12, height = 14)

Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)
#ggsave('test2.pdf', plot = Fig2, units = 'in', width = 14, height = 12)
